Comparative effectiveness of ChAdOx1 versus BNT162b2 COVID-19 vaccines in Health and Social Care workers in England: a cohort study using OpenSAFELY

### 2022-03-30 manuscript/draft-manuscript-v1-1.Rmd
some revisions

*Setting* Linked primary care, hospital, and COVID-19 surveillance records available within the OpenSAFELY-TPP research platform.

@@

*Main outcome measures* Recorded SARS-CoV-2 positive test, or COVID-19 related Accident and Emergency attendance or hospital admission occurring within 20 weeks of vaccination.

*Results* The cumulative incidence of each outcome was similar for both vaccines during the first 20 weeks post-vaccination. The cumulative incidence of recorded SARS-CoV-2 infection 6 weeks after vaccination with BNT162b2 was `r filter(cmlinc_ns3, outcome=="postest", vax1_az==0, tstop==7*6)$inc_format` per 1000 people (95%CI `r filter(cmlinc_ns3, outcome=="postest", vax1_az==0, tstop==7*6)$inc_CI_to`) and with ChAdOx1 was `r filter(cmlinc_ns3, outcome=="postest", vax1_az==1, tstop==7*6)$inc_format` (95%CI `r filter(cmlinc_ns3, outcome=="postest", vax1_az==1, tstop==7*6)$inc_CI_to`), representing a difference of `r filter(riskdiff_ns3, outcome=="postest", tstop==7*6)$diff_format` per 1000 people (95%CI `r filter(riskdiff_ns3, outcome=="postest", tstop==7*6)$diff_CI_to`). The difference in the cumulative incidence per 1000 people of COVID-19 accident and emergency attendance at 6 weeks was `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*6)$diff_format` per 1000 people (95%CI `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*6)$diff_CI_to`). For COVID-19 hospital admission, this difference was `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*6)$diff_format` per 1000 people (95%CI `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*6)$diff_CI_to`).

*Conclusions* In this cohort of healthcare workers where we would not anticipate vaccine type to be related to health status, we found no substantial differences in the incidence of SARS-CoV-2 infection or COVID-19 disease up to 20 weeks after vaccination. Incidence dropped sharply after 3-4 weeks and there were very few COVID-19 hospital attendance and admission events after this period. This is in line with expected onset of vaccine-induced immunity, and suggests strong protection against COVID-19 disease for both vaccines.

@@

The COVID-19 global pandemic has prompted the rapid development and delivery of vaccines to combat the disease. Following demonstration of high safety and efficacy against symptomatic and severe disease in phase-III randomised controlled trials (RCTs), two vaccines have been approved and widely administered as part of the national vaccination programme in the United Kingdom: the Pfizer-BioNTech BNT162b2 mRNA COVID-19 vaccine @polack2020 (*BNT162b2*) and the Oxford-AstraZeneca ChAdOx1 nCoV-19 vaccine @voysey2021 (*ChAdOx1*). Post-authorisation assessment of vaccine effectiveness using observational data is necessary to monitor the success of such programmes as, invariably, target populations and settings differ substantially from those of trials. To date, there have been no RCTs that have directly compared the BNT162b2 and ChAdOx1 vaccines to estimate the relative efficacy against COVID-19 infection and disease in the same population. The concurrent roll-out of these vaccines across the UK @collaborative2021, combined with the country's well-developed infrastructure for conducting research using routinely-collected primary care health data, provides a rare opportunity to emulate such a trial using observational data.

COVID-19 vaccination in the UK has been prioritised based on the risk of infection and subsequent severity of disease @covid19. Health and social care workers (HCWs) were amongst the first groups eligible for vaccination due to the high occupational exposure to the SARS-CoV-2 virus, making HCW status an important confounder of the effect of vaccination on infection and post-infection disease. By studying this group in isolation, this confounding is mitigated, and vaccine effects in a healthy, high-exposure group can be estimated.

This cohort study used a target-trial design to assess the effectiveness of ChAdOx1 compared with BNT162b2 in HCWs, including second dose effectiveness, using the OpenSAFELY-TPP linked primary care database covering around 40% of England's population.

@@

The OpenSAFELY-TPP database (<https://opensafely.org>) covers 24 million people registered at GP practices that use TPP SystmOne electronic health record software. It includes pseudonymized data such as coded diagnoses, medications and physiological parameters. No free text data are included. This primary care data is linked (via NHS numbers) with A&E attendance and in-patient hospital spell records via NHS Digital's Hospital Episode Statistics (HES), national coronavirus testing records via the Second Generation Surveillance System (SGSS), and national death registry records from the Office for National Statistics (ONS). Vaccination status is available in the GP record directly via the National Immunisation Management System (NIMS). HCW status is recorded for all vaccine recipients at the time of vaccination, and this information is sent to OpenSAFELY-TPP from NHS Digital's COVID-19 data store.

@@

We studied health and social care workers in England vaccinated with either BNT162b2 or ChAdOx1. This group was prioritised for vaccination at the start of the vaccine roll-out due to the high occupational exposure to the SARS-CoV-2 virus, and many were vaccinated during the period where both vaccines were widely used.

@@

Three outcomes were defined: positive SARS-CoV-2 test; COVID-19 A&E attendance; and unplanned COVID-19 hospital admission. Positive SARS-CoV-2 tests were identified using SGSS records and based on swab date. Both polymerase chain reaction (PCR) and lateral flow tests are included, without differentiation between symptomatic and asymptomatic infection. COVID-19 A&E attendances were identified using HES emergency care records with U07.1 ("COVID-19, virus identified") or U07.2 ("COVID-19, virus not identified") ICD-10 diagnosis codes @emergenc . Unplanned COVID-19 hospital admissions were identified using HES in-patient hospital records with U07.1 or U07.2 reason for admission ICD-10 codes.

@@

At 12 weeks, `r label_percent(0.1)(baseline_pfizer$seconddose_12week)` of BNT162b2 recipients had received a second dose, compared with `r label_percent(0.1)(baseline_az$seconddose_12week)` of ChAdOx1 recipients. There were `r label_number(1)(crossbrand_n)` (`r label_percent(0.01)(crossbrand_prop)`) participants who received a second dose within 20 weeks that was not the same brand as the first dose, including Moderna mRNA-1273 vaccine second doses (supplementary Figure S3, Table S4).

@@

At 6 weeks post-vaccination, the ChAdOx1 versus BNT162b2 absolute risk difference per 1000 people (Figure 2) for a positive SARS-CoV-2 test was `r filter(riskdiff_ns3, outcome=="postest", tstop==7*6)$diff_format` (95%CI `r filter(riskdiff_ns3, outcome=="postest", tstop==7*6)$diff_CI_to`), for COVID-19 A&E attendances was `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*6)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*6)$diff_CI_to`) and for COVID-19 hospital admissions was `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*6)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*6)$diff_CI_to`).

At 18 weeks post-vaccination, the absolute risk difference per 1000 people for a positive SARS-CoV-2 test was `r filter(riskdiff_ns3, outcome=="postest", tstop==7*18)$diff_format` (95%CI `r filter(riskdiff_ns3, outcome=="postest", tstop==7*18)$diff_CI_to`), for COVID-19 A&E attendances was `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*18)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*18)$diff_CI_to`) and for COVID-19 hospital admissions was `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*18)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*18)$diff_CI_to`).

The estimated log hazard ratios were non-zero at certain periods (Figure 2), but the underlying absolute risk was small enough that this did not manifest in substantial differences in the cumulative incidence beyond the fourth week.

@@

  plot_riskdiff_ns <-
    riskdiff_ns3 %>%
    ggplot()+
    geom_hline(aes(yintercept=0), colour='black')+
    geom_step(aes(x=tstop, y=diff*1000), colour='orchid4')+
    geom_rect(aes(xmin=tstop, xmax=lead_tstop, ymin=diff.ll*1000, ymax=diff.ul*1000), alpha=0.1, colour="transparent", fill='orchid4')+
    facet_wrap(facets=vars(outcome_descr), nrow=1, scales="free_y")+
    scale_x_continuous(
      limits = c(0, NA),
      breaks = seq(0,7*52,by=28),
      expand = expansion(0)
    )+
    scale_y_continuous(
      expand = expansion(0),
      sec.axis = dup_axis(name="Higher incidence after \n <- ChAdOx1 / BNT162b2 ->", breaks = NULL)
    )+
    coord_cartesian(ylim= c(-3, 3))+
    scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
    scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
    labs(
      x=NULL,
      y="Difference\nin cumul. incidence\nper 1000",
      colour=NULL
    )+
    theme_minimal(base_size = 9)+
    theme(
      axis.line.y.left = element_line(colour = "black"),

@@

  ## hazard ratio
  plot_hazard_ns <-
    hazard_ns3 %>%
    ggplot()+
    geom_hline(aes(yintercept=1), colour='black')+
    geom_step(aes(x=tstop, y=hr), colour='orchid4', direction="vh")+
    geom_rect(aes(xmin=lag_tstop, xmax=tstop, ymin=hr.ll, ymax=hr.ul), alpha=0.1, colour="transparent", fill='orchid4')+
    facet_wrap(facets=vars(outcome_descr), nrow=1, scales="free_y")+
    scale_x_continuous(
      limits = c(0, NA),
      breaks = seq(0,7*52,by=28),
      expand = expansion(0)
    )+
    scale_y_log10(
      breaks=c(0.1, 0.2, 0.5, 1, 2, 5, 10),
      sec.axis = dup_axis(name="Higher hazard after \n <- ChAdOx1 / BNT162b2 ->", breaks = NULL)
    )+
    coord_cartesian(ylim= c(0.1, 10))+
    labs(
      x = "Days since first dose",
      y = "Hazard ratio"
    )+
    theme_minimal(base_size = 9)+
    theme(
      axis.line.y.left = element_line(colour = "black"),

@@

This observational study of `r label_number(1, big.mark=",")(flowchart[[nrow(flowchart), "n"]])` health and social care workers living in England found similar outcomes for those receiving the BNT162b2 or ChAdOx1 COVID-19 vaccines during the first 20 weeks post-vaccination. There is a clear leveling-off of the incidence of positive tests after around 3-4 weeks for both vaccines, consistent with the expected time-to-onset of vaccine-induced immunity of around 2 weeks plus the delay from infection to a positive test. At 20 weeks, the difference in the cumulative incidence of positive test was `r filter(riskdiff_ns3, outcome=="postest", tstop==7*20)$diff_format` per 1000 people `r filter(riskdiff_ns3, outcome=="postest", tstop==7*20)$diff_CI` higher for ChAdOx1 compared with BNT162b2, and extrapolating beyond this time would suggest a small advantage for BNT162b2. However, BNT162b2 recipients were more likely to receive their second dose sooner, conferring a greater protective effect that is not accounted for in this estimate.

There were very few COVID-19 hospital attendance and admission events after this period, and the difference in the cumulative incidence (and their CIs) was well below 1 event per 1000 people in either direction.

@@

Despite reasonable balance of vaccine allocation across baseline characteristics and adjustment for a range of potential confounders, the possibility of unmeasured confounding remains. The cold storage requirements of BNT162b2 meant that it was more likely to have be administered in Acute NHS Trusts and other large vaccination centres, and is thus a potential confounder for instance due to differences in viral exposure across these settings. Though we adjusted for region, rurality, and deprivation, we were unable to directly account for occupational differences that may affect both exposure risk and vaccine type. We note also that the GP practices covered in this study are not geographically representative, as TPP SystmOne software is not widely-used in some regions such as London and the North West.

@@

Second-dose effects were considered using an intention-to-treat approach, which does not account for potential differences in the timing of the second dose between vaccine types. However, we found that by 12 weeks more BNT162b2 recipients had received a second dose than ChAdOx1 recipients, which would bias effect estimates in favour of BNT162b2 assuming higher protection after a second-dose of either vaccine. Regardless, there was insufficient power to reliably estimate comparative second-dose effects as event incidence at 12 weeks and beyond was low for both vaccines, with fewer than 10 positive tests per 1000-person-years, and far fewer for COVID-19 A&E attendances and hospital admissions during this period. The need for many thousands more person-years to detect differences in second-dose effects is encouraging to the extent that it suggests extremely high protective effects for both vaccines, though the fact that background case rates in England were declining over the duration of the study period should not be ignored.

@@

A number of studies have estimated COVID-19 vaccine effectiveness in observational data @vasileiou2021 @cabezas2021 @effectiv @hall2021. Vaccine protection is typically assessed by comparing vaccinated against unvaccinated person-time, or against the early post-vaccination period before anticipated onset of vaccine-induced immunity. Such designs are vulnerable to confounding and selection bias, for example due to vaccine prioritisation and eligibility policies. In addition, differences in vaccine access and acceptance can cause substantial imbalance between vaccinated and unvaccinated groups that may be impossible to account for. Many of these biases can be bypassed when making direct comparisons between recipients of different vaccine types, if the relative performance of two vaccines is of interest. To our knowledge, the present study is the first to assess effectiveness of the BNT162b2 and ChAdOx1 vaccines in a head-to-head comparison in either an experimental or observational setting.

Our study suggests that there is little to differentiate the vaccines with respect to their respective protective effects against infection and hospital-related COVID-19 outcomes. This pushes decisions about which vaccine to favour onto other attributes such as safety profile, logistics, and cost. However, we cannot rule-out larger differences in effectiveness in other settings. For instance: results seen in this cohort of predominantly healthy HCWs may not reflect comparative effectiveness in more vulnerable groups, such as the elderly or immunosuppressed; the dominant circulating variant during the period of study was the Alpha variant but has now been replaced by Delta variant against which vaccines may be less effective @bernal2021; longer-term effectiveness, and in particular the potential immunological waning, may differ.

@@

This study found no substantial differences in the incidence of SARS-CoV-2 infection or COVID-19-related hospital events following vaccination with BNT162b2 or ChAdOx1 in a cohort of health and social care workers in England. The incidence of COVID-19-related A&E attendances or hospital admissions four weeks after the first dose of either vaccine was fewer than one event in 10,000-person-years, suggesting strong protective effects for both vaccines but nevertheless limiting power to reliably estimate comparative effectiveness with respect to severe COVID-19 outcomes. Further studies are needed to assess comparative effectiveness in newer, more prevalent variants and to assess longer-term effectiveness.

>>>>

*Setting* Linked primary care, hospital, and COVID-19 surveillance records available within the OpenSAFELY-TPP research platform, covering a period where the SARS-CoV-2 Alpha variant was dominant.

@@

*Main outcome measures* Recorded SARS-CoV-2 positive test, or COVID-19 related Accident and Emergency attendance or hospital admission occurring within 20 weeks of receipt of the first vaccine dose.

*Results* The cumulative incidence of each outcome was similar for both vaccines during the first 20 weeks post-vaccination. The cumulative incidence of recorded SARS-CoV-2 infection 20 weeks after first-dose vaccination with BNT162b2 was `r filter(cmlinc_ns3, outcome=="postest", vax1_az==0, tstop==7*20)$inc_format` per 1000 people (95%CI `r filter(cmlinc_ns3, outcome=="postest", vax1_az==0, tstop==7*20)$inc_CI_to`) and with ChAdOx1 was `r filter(cmlinc_ns3, outcome=="postest", vax1_az==1, tstop==7*20)$inc_format` (95%CI `r filter(cmlinc_ns3, outcome=="postest", vax1_az==1, tstop==7*20)$inc_CI_to`), representing a difference of `r filter(riskdiff_ns3, outcome=="postest", tstop==7*20)$diff_format` per 1000 people (95%CI `r filter(riskdiff_ns3, outcome=="postest", tstop==7*20)$diff_CI_to`). The difference in the cumulative incidence per 1000 people of COVID-19 accident and emergency attendance at 20 weeks was `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*20)$diff_format` per 1000 people (95%CI `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*20)$diff_CI_to`). For COVID-19 hospital admission, this difference was `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*20)$diff_format` per 1000 people (95%CI `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*20)$diff_CI_to`).

*Conclusions* In this cohort of healthcare workers where we would not anticipate vaccine type to be related to health status, we found no substantial differences in the incidence of SARS-CoV-2 infection or COVID-19 disease up to 20 weeks after vaccination. Incidence dropped sharply at 3-4 weeks post-vaccination and there were very few COVID-19 hospital attendance and admission events after this period. This is in line with expected onset of vaccine-induced immunity, and suggests strong protection against Alpha-variant COVID-19 disease for both vaccines in this relatively young and healthy population of HCWs.

@@

The COVID-19 global pandemic has prompted the rapid development and delivery of vaccines to combat the disease. Following demonstration of high safety and efficacy against symptomatic and severe disease in phase-III randomised controlled trials (RCTs), two vaccines have been approved and widely administered as part of the national vaccination programme in the United Kingdom: the Pfizer-BioNTech BNT162b2 mRNA COVID-19 vaccine @polack2020 (*BNT162b2*) and the Oxford-AstraZeneca ChAdOx1 nCoV-19 viral vector vaccine @voysey2021 (*ChAdOx1*). Post-authorisation assessment of vaccine effectiveness using observational data is necessary to monitor the success of such programmes as, invariably, target populations and settings differ substantially from those of trials. To date, there have been no RCTs that have directly compared the BNT162b2 and ChAdOx1 vaccines to estimate the relative efficacy against COVID-19 infection and disease in the same population. The concurrent roll-out of these vaccines across the UK @collaborative2021, combined with the country's well-developed infrastructure for conducting research using routinely-collected primary care health data, provides a rare opportunity to emulate such a trial using observational data.

COVID-19 vaccination in the UK has been prioritised based on the risk of infection and subsequent severity of disease @covid19. Health and social care workers (HCWs) were amongst the first groups eligible for vaccination due to the potentially high occupational exposure to the SARS-CoV-2 virus, making HCW status an important confounder of the effect of vaccination on infection and post-infection disease. By studying this group in isolation, this confounding is mitigated. Further, unlike many other young people vaccinated later, HCWs were vaccinated during a period of high community infection, increasing event rates and therefore statistical power.

This cohort study used a target-trial design to assess the effectiveness of ChAdOx1 compared with BNT162b2 against COVID-19-related outcomes in HCWs, including second dose effectiveness, using the OpenSAFELY-TPP linked primary care database covering around 40% of England's population. Effectiveness was assessed primarily in an era where the SARS-CoV-2 Alpha variant was dominant.

@@

The OpenSAFELY-TPP database (<https://opensafely.org>) covers 24 million people registered at National Health Service (NHS) General Practices using TPP SystmOne electronic health record software. It includes pseudonymized data such as coded diagnoses, medications and physiological parameters. No free text data are included. This primary care data is linked, via NHS numbers, with Accident & Emergency (A&E) attendance and in-patient hospital spell records via NHS Digital's Hospital Episode Statistics (HES), national coronavirus testing records via the Second Generation Surveillance System (SGSS), and national death registry records from the Office for National Statistics (ONS). Vaccination status is available in the GP record directly via the National Immunisation Management System (NIMS). Self-designated HCW status is recorded for all vaccine recipients at the time of vaccination, and this information is sent to OpenSAFELY-TPP from NHS Digital's COVID-19 data store.

@@

We studied health and social care workers in England vaccinated with either BNT162b2 or ChAdOx1. This group was prioritised for vaccination at the start of the vaccine roll-out due to the potentially high occupational exposure to the SARS-CoV-2 virus, and many were vaccinated during the period where both vaccines were widely used.

@@

Three outcomes were defined: positive SARS-CoV-2 test; COVID-19 A&E attendance; and unplanned COVID-19 hospital admission. Positive SARS-CoV-2 tests were identified using SGSS records and based on swab date. Both polymerase chain reaction (PCR) and lateral flow positive tests were included, without differentiation between symptomatic and asymptomatic infection. Test provenance, for example patient-initiated tests versus employee-screening, was not known. COVID-19 A&E attendances were identified using HES emergency care records with U07.1 ("COVID-19, virus identified") or U07.2 ("COVID-19, virus not identified") ICD-10 diagnosis codes @emergenc . Unplanned COVID-19 hospital admissions were identified using HES in-patient hospital records with U07.1 or U07.2 reason for admission ICD-10 codes.

@@

## Second doses

At 12 weeks (72 days), `r label_percent(0.1)(baseline_pfizer$seconddose_12week)` of BNT162b2 recipients had received a second dose, compared with `r label_percent(0.1)(baseline_az$seconddose_12week)` of ChAdOx1 recipients. There were `r label_number(1)(crossbrand_n)` (`r label_percent(0.01)(crossbrand_prop)`) participants who received a second dose within 20 weeks (140 days) that was not the same brand as the first dose, including Moderna mRNA-1273 vaccine second doses (supplementary Figure S3, Table S4).

@@

By 6 weeks (42 days) post-vaccination, prior to receipt of the second dose for the vast majority of participants, the ChAdOx1 versus BNT162b2 absolute risk difference per 1000 people (Figure 2) for a positive SARS-CoV-2 test was `r filter(riskdiff_ns3, outcome=="postest", tstop==7*6)$diff_format` (95%CI `r filter(riskdiff_ns3, outcome=="postest", tstop==7*6)$diff_CI_to`), for COVID-19 A&E attendances was `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*6)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*6)$diff_CI_to`) and for COVID-19 hospital admissions was `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*6)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*6)$diff_CI_to`).

At 20 weeks (140 days) post-vaccination, at which time most participants had received their second dose, the absolute risk difference per 1000 people for a positive SARS-CoV-2 test was `r filter(riskdiff_ns3, outcome=="postest", tstop==7*20)$diff_format` (95%CI `r filter(riskdiff_ns3, outcome=="postest", tstop==7*20)$diff_CI_to`), for COVID-19 A&E attendances was `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*20)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*20)$diff_CI_to`) and for COVID-19 hospital admissions was `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*20)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*20)$diff_CI_to`). We note that extrapolating beyond 20 weeks for positive SARS-CoV-2 tests, we would infer a small advantage for BNT162b2 in terms of absolute risk.

In general, events rates are much higher in the first few weeks after vaccination, and then there is a clear leveling-off after around 3-4 weeks for both vaccines, beyond which the event rates were very low. For COVID-19 hospital attendance and admission events in particular, the difference in the cumulative incidence (and their CIs) was well below 1 event per 1000 people in either direction. The estimated log hazard ratios were non-zero at certain periods (Figure 2), but the underlying absolute risk was small enough that this did not manifest in substantial differences in the cumulative incidence beyond the fourth week.

@@

  plot_riskdiff_ns <-
    riskdiff_ns3 %>%
    ggplot()+
    geom_hline(aes(yintercept=0), colour='black')+
    geom_step(aes(x=tstop, y=diff*1000), colour='orchid4')+
    geom_rect(aes(xmin=tstop, xmax=lead_tstop, ymin=diff.ll*1000, ymax=diff.ul*1000), alpha=0.1, colour="transparent", fill='orchid4')+
    facet_wrap(facets=vars(outcome_descr), nrow=1, scales="free_y")+
    scale_x_continuous(
      limits = c(0, NA),
      breaks = seq(0,7*52,by=28),
      expand = expansion(0)
    )+
    scale_y_continuous(
      expand = expansion(0),
      sec.axis = dup_axis(name="Higher incidence after \n <- ChAdOx1 | BNT162b2 ->", breaks = NULL)
    )+
    coord_cartesian(ylim= c(-3, 3))+
    scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
    scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
    labs(
      x=NULL,
      y="Difference\nin cumul. incidence\nper 1000",
      colour=NULL
    )+
    theme_minimal(base_size = 9)+
    theme(
      axis.line.y.left = element_line(colour = "black"),

@@

  ## hazard ratio
  plot_hazard_ns <-
    hazard_ns3 %>%
    ggplot()+
    geom_hline(aes(yintercept=1), colour='black')+
    geom_step(aes(x=tstop, y=hr), colour='orchid4', direction="vh")+
    geom_rect(aes(xmin=lag_tstop, xmax=tstop, ymin=hr.ll, ymax=hr.ul), alpha=0.1, colour="transparent", fill='orchid4')+
    facet_wrap(facets=vars(outcome_descr), nrow=1, scales="free_y")+
    scale_x_continuous(
      limits = c(0, NA),
      breaks = seq(0,7*52,by=28),
      expand = expansion(0)
    )+
    scale_y_log10(
      breaks=c(0.1, 0.2, 0.5, 1, 2, 5, 10),
      sec.axis = dup_axis(name="Higher hazard after \n <- ChAdOx1 | BNT162b2 ->", breaks = NULL)
    )+
    coord_cartesian(ylim= c(0.1, 10))+
    labs(
      x = "Days since first dose",
      y = "Hazard ratio"
    )+
    theme_minimal(base_size = 9)+
    theme(
      axis.line.y.left = element_line(colour = "black"),

@@

## Key findings

This observational study of `r label_number(1, big.mark=",")(flowchart[[nrow(flowchart), "n"]])` adult health and social care workers living in England found similar outcomes for those receiving the BNT162b2 or ChAdOx1 COVID-19 vaccines during the first 20 weeks post-vaccination in the Alpha variant era. Consistent with the expected time-to-onset of vaccine-induced immunity at around 2 weeks post-vaccination, plus the delay from infection to the event of interest, event rates tapered after the first few weeks. There were very few COVID-19 hospital attendance and admission events after this period, suggesting strong protective effects for both vaccines but nevertheless limiting power to reliably estimate comparative effectiveness with respect to severe COVID-19 outcomes.

There is some evidence that, compared with ChAdOx1 recipients, there was better longer-term protection for BNT162b2 recipients, but they were more also likely to receive their second dose sooner, likely conferring a greater protective effect that is not accounted for in the intention-to-treat effectiveness estimates.

@@

Despite reasonable balance of vaccine allocation across baseline characteristics and adjustment for a range of potential confounders, the possibility of unmeasured confounding remains. The cold storage requirements of BNT162b2 meant that it was more likely to have be administered in Acute NHS Trusts and other large vaccination centres, and is thus a potential confounder for instance due to differences in viral exposure across these settings. Though we adjusted for region, rurality, and deprivation, we were unable to directly account for occupational differences that may affect both exposure risk and vaccine type. For instance, we had no means to quantify any differences in HCW specialty, working hours, or patient contact time between the treatment groups. The primary care practices covered in this study are not geographically representative, as TPP SystmOne software is not widely-used in some regions such as London and the North West.

@@

Second-dose effects were considered using an intention-to-treat approach, which does not account for potential differences in the timing of the second dose between vaccine types. However, we found that by 12 weeks more BNT162b2 recipients had received a second dose than ChAdOx1 recipients, which would bias effect estimates in favour of BNT162b2 assuming higher protection after a second-dose of either vaccine. Regardless, there was insufficient power to reliably estimate comparative second-dose effects as event incidence at 12 weeks and beyond was low for both vaccines, with fewer than 10 positive tests per 1000-person-years, and far fewer for COVID-19 A&E attendances and hospital admissions during this period. The need for many thousands more person-years to detect differences in second-dose effects is encouraging to the extent that it suggests extremely high protective effects for both vaccines, though declining background case rates in England over the duration of the study period (Figure S4) would have also contributed to reduced event rates in both treatment groups.

@@

A number of studies have estimated first-dose COVID-19 vaccine effectiveness in observational data @Iheanacho2021, including BNT162b2 and ChAdOx1 vaccines in similar populations @vasileiou2021 @cabezas2021 @effectiv @bernal2021 @Shrotri2021 and in healthcare workers @Paris2021 @hall2021. Vaccine protection is typically assessed by comparing vaccinated against unvaccinated person-time, or against the early post-vaccination period before anticipated onset of vaccine-induced immunity. Such designs are vulnerable to confounding and selection bias, for example due to vaccine prioritisation and eligibility policies and risk or health-seeking behaviours. In addition, differences in vaccine access and acceptance can cause substantial imbalance between vaccinated and unvaccinated groups that may be impossible to account for. Many of these biases can be bypassed when making direct comparisons between recipients of different vaccine types, if the relative performance of two vaccines is of interest \@@Dickerman2022. To our knowledge, the present study is the first to assess effectiveness of the BNT162b2 and ChAdOx1 vaccines in a head-to-head comparison in either an experimental or observational setting.

Our study suggests that there is little to differentiate the vaccines with respect to their protective effects against infection and hospital-related COVID-19 outcomes within the first 20 weeks after the first dose in adult HCWs under 65 years old. This pushes decisions about which vaccine to favour onto other attributes such as safety profile, logistics, and cost. However, we cannot rule-out larger differences in effectiveness in other settings. For instance: results seen in this cohort of predominantly healthy HCWs may not reflect comparative effectiveness in more vulnerable groups, such as the elderly or immunosuppressed; the dominant circulating variant during the period of study was the Alpha variant but has now been replaced by Delta variant against which vaccines may be less effective @bernal2021; longer-term effectiveness, and in particular the potential immunological waning, may differ.

@@

This study found no substantial differences in the incidence of SARS-CoV-2 infection or COVID-19-related hospital events following vaccination with BNT162b2 or ChAdOx1 in a cohort of health and social care workers in England in the Alpha era. There were very few severe COVID-19 outcomes after the expected onset of vaccine-induced immunity at around 2 weeks, reducing power to reliably compare effectiveness against these outcomes. Further studies are needed to assess comparative effectiveness against newer, more prevalent variants and to assess longer-term effectiveness.
